### 2023-09-01 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@514dee0640609a4566841d9b3736fc708e991cad 🚀

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{\textbf{Plowright2017PatZoo?}} Although host
richness is often used as a superficial proxy for spillover
risk,\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?},\textbf{Sanchez2022Strategy?}}
these approaches oversimplify the relevant interspecific heterogeneity
in immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{\textbf{Agosta2010HowSpe?}} In the case of
generalist pathogens like betacoronaviruses, there is conceptual and
empirical support to the idea that these community-level mechanisms are
even more important,\textsuperscript{\textbf{Power2004PatSpi?}}
particularly given that cross-species transmission may, as a rule,
structure viral evolution more than co-divergence with
hosts.\textsuperscript{\textbf{Geoghegan2017ComAna?}} This creates a
disconnect between coevolutionary theory and most existing ecological
frameworks for mapping spillover risk.

@@

\begin{figure}
\hypertarget{fig:biogeo}{%
\centering
\includegraphics{figures/combined_biogeo.png}
\caption{\textbf{Bat and betacoronavirus biogeographic regions.}
Phylogeography of bats (top) and viruses (bottom) is categorized based
on analysis of bat distributions paired with bat or virus phylogeny. The
different colors show tendencies to separate alongside the first two
components of a PCoA. Note that the PCoA for the bats and viruses are
independent, and so cannot be compared directly -- that being said, the
regions can be compared across maps.}\label{fig:biogeo}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:richness}{%
\centering
\includegraphics{figures/combined_richness.png}
\caption{\textbf{Bat and betacoronavirus diversity.} Top panel: relative
diversity of known bat hosts of betacoronaviruses. This map shows that
the region with the largest number of possible hosts is South-Eastern
Asia. Bottom panel: congruence between the evolutionary distinctiveness
of the hosts (grey to blue) and the viruses (grey to
red).}\label{fig:richness}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:trivariate}{%
\centering
\includegraphics{figures/risk_trivariate.png}
\caption{\textbf{Trivariate additive mapping of the components of risk.}
Viral sharing runs from yellow (low) to blue (high); host phylogenetic
diversity runs from pink (low) to high (green); and host compositional
uniqueness runs from cyan (low) to red (high). The GMTC suggests that
the highest evolutionary potential for emergence exists in unique and
diverse host communities with low viral sharing, \emph{i.e.} pixels
around yellow. All components within bat host ranges are scaled in
brightness so that a pixel with no sharing, no phylogenetic diversity,
and no compositional uniqueness would be black, and a pixel with maximal
values for each would be white.}\label{fig:trivariate}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:risk}{%
\centering
\includegraphics{figures/risk_map.png}
\caption{\textbf{Evolutionary potential for zoonotic emergence of
bat-origin betacoronaviruses.} Risk is a composite measure of the color
value and angular distance to the yellow hue in
fig.~\ref{fig:trivariate} (see Methods).}\label{fig:risk}
}
\end{figure}

@@

Bats---and the spillover of their viruses---are also sensitive to
anthropogenic factors others than climate change, including
deforestation and other kinds of habitat loss, increased stress, and
greater contact with potential bridge hosts like domesticated
species.\textsuperscript{\textbf{Alves2018GeoVar?},\textbf{Treitler2016EffLoc?},\textbf{Rulli2021LanCha?},\textbf{Mendenhall2014PreBio?}}
This represents a challenge for both conservation strategies and
pandemic prevention,\textsuperscript{\textbf{Amman2011InvRol?}} but
identifying areas at risk, and protecting the health of bats and
ecosystems within those zones, can be a win-win intervention for
both.\textsuperscript{\textbf{Hopkins2021HowIde?},\textbf{Plowright2021LanUse?},\textbf{OHHLEP2022OneHea?}}
As we scale these predictions down in space to finer spatial resolutions
to guide public health actions,\textsuperscript{1} the incorporation of
human activity predictors will become more
importyant.\textsuperscript{2}

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{\textbf{Albery2020PreGlo?}} This network stores
pairwise values of viral community similarity, measured for all hosts
(to maintain consistency with teh phylogenetic diversity measure) across
all viruses; therefore, we consider that it accounts for some overall
similarity in the way hosts deal with viruses, and not only
betacoronaviruses. There is empirical evidence that capacity for
cross-species transmission even between divergent species is generally
high,\textsuperscript{3} especially for
beta-coronaviruses.\textsuperscript{4} To project viral sharing values
into a single value for every pixel, we averaged the pairwise scores.
High values of the average sharing propensity means that this specific
extant bat assemblage is likely to be proficient at exchanging viruses.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-Muylaert2022Present}{}%
\CSLLeftMargin{1. }
\CSLRightInline{Muylaert, R. L. \emph{et al.} Present and future
distribution of bat hosts of sarbecoviruses: Implications for
conservation and public health. \emph{Proceedings of the Royal Society
B: Biological Sciences} \textbf{289}, 20220397 (2022).}

\leavevmode\hypertarget{ref-Ka-WaiHui2006Reasons}{}%
\CSLLeftMargin{2. }
\CSLRightInline{Ka-Wai Hui, E. Reasons for the increase in emerging and
re-emerging viral infectious diseases. \emph{Microbes and Infection}
\textbf{8}, 905--916 (2006).}

\leavevmode\hypertarget{ref-Mollentze2020Viral}{}%
\CSLLeftMargin{3. }
\CSLRightInline{Mollentze, N. \& Streicker, D. G. Viral zoonotic risk is
homogenous among taxonomic orders of mammalian and avian reservoir
hosts. \emph{Proceedings of the National Academy of Sciences}
\textbf{117}, 9423--9430 (2020).}

\leavevmode\hypertarget{ref-Latinne2020Origin}{}%
\CSLLeftMargin{4. }
\CSLRightInline{Latinne, A. \emph{et al.} Origin and cross-species
transmission of bat coronaviruses in China. \emph{bioRxiv: The Preprint
Server for Biology} 2020.05.31.116061 (2020)
doi:\href{https://doi.org/10.1101/2020.05.31.116061}{10.1101/2020.05.31.116061}.}

>>>>

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{1} Although host richness is often used as a
superficial proxy for spillover
risk,\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?},\textbf{Sanchez2022Strategy?}}
these approaches oversimplify the relevant interspecific heterogeneity
in immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{2} In the case of generalist pathogens like
betacoronaviruses, there is conceptual and empirical support to the idea
that these community-level mechanisms are even more
important,\textsuperscript{3} particularly given that cross-species
transmission may, as a rule, structure viral evolution more than
co-divergence with hosts.\textsuperscript{4} This creates a disconnect
between coevolutionary theory and most existing ecological frameworks
for mapping spillover risk.

@@

\begin{figure}
\hypertarget{fig:biogeo}{%
\centering
\includegraphics{figures/combined_biogeo.png}
\caption{\textbf{Bat and betacoronavirus biogeographic regions.}
Phylogeography of bats (top) and viruses (bottom) is categorized based
on an analysis of bat distributions, paired with bat or virus phylogeny.
The different colors show tendencies to separate alongside the first two
components of a PCoA. Note that the PCoA for the bats and viruses are
independent, and so cannot be compared directly -- that being said, the
fact that different regions cluster in the same way across maps be
directly compared.}\label{fig:biogeo}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:richness}{%
\centering
\includegraphics{figures/combined_richness.png}
\caption{\textbf{Bat and betacoronavirus diversity.} Top panel:
diversity of known bat hosts of betacoronaviruses in our dataset. This
map shows that the region with the largest number of possible hosts is
South-Eastern Asia. Bottom panel: congruence between the
\emph{evolutionary} distinctiveness of the hosts (grey to blue) and the
viruses (grey to red). Darker areas have higher combined evolutionary
distinctiveness for the entire bat-virus system.}\label{fig:richness}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:trivariate}{%
\centering
\includegraphics{figures/risk_trivariate.png}
\caption{\textbf{Trivariate additive mapping of the components of risk.}
Viral sharing runs from yellow (low) to blue (high); host phylogenetic
diversity runs from pink (low) to high (green); and host compositional
uniqueness runs from cyan (low) to red (high). The GMTC suggests that
the highest evolutionary potential for emergence exists in unique and
diverse host communities with low viral sharing, \emph{i.e.} pixels
around yellow. All components within bat host ranges are scaled in
brightness so that a pixel with no sharing, no phylogenetic diversity,
and no compositional uniqueness would be black, and a pixel with maximal
values for each would be white. The individual layers that compose this
figure are given in supplementary material.}\label{fig:trivariate}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:risk}{%
\centering
\includegraphics{figures/risk_map.png}
\caption{\textbf{Evolutionary potential for zoonotic emergence of
bat-origin betacoronaviruses.} Risk is a composite measure of the color
value and angular distance to the yellow hue in
fig.~\ref{fig:trivariate} (see Methods). Darker pixels represent areas
where the co-evolutionary mechanisms are likely to introduce a strong
risk of emergence.}\label{fig:risk}
}
\end{figure}

@@

Bats---and the spillover of their viruses---are also sensitive to
anthropogenic factors others than climate change, including
deforestation and other kinds of habitat loss, increased stress, and
greater contact with potential bridge hosts like domesticated
species.\textsuperscript{\textbf{Alves2018GeoVar?},\textbf{Treitler2016EffLoc?},\textbf{Rulli2021LanCha?},\textbf{Mendenhall2014PreBio?}}
This represents a challenge for both conservation strategies and
pandemic prevention,\textsuperscript{\textbf{Amman2011InvRol?}} but
identifying areas at risk, and protecting the health of bats and
ecosystems within those zones, can be a win-win intervention for
both.\textsuperscript{\textbf{Hopkins2021HowIde?},\textbf{Plowright2021LanUse?},\textbf{OHHLEP2022OneHea?}}
As we scale these predictions down in space to finer spatial resolutions
to guide public health actions,\textsuperscript{5} the incorporation of
human activity predictors will become more
importyant.\textsuperscript{6}

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{\textbf{Albery2020PreGlo?}} This network stores
pairwise values of viral community similarity, measured for all hosts
(to maintain consistency with teh phylogenetic diversity measure) across
all viruses; therefore, we consider that it accounts for some overall
similarity in the way hosts deal with viruses, and not only
betacoronaviruses. There is empirical evidence that capacity for
cross-species transmission even between divergent species is generally
high,\textsuperscript{7} especially for
beta-coronaviruses.\textsuperscript{8} To project viral sharing values
into a single value for every pixel, we averaged the pairwise scores.
High values of the average sharing propensity means that this specific
extant bat assemblage is likely to be proficient at exchanging viruses.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-Plowright2017Pathways}{}%
\CSLLeftMargin{1. }
\CSLRightInline{Plowright, R. K. \emph{et al.} Pathways to zoonotic
spillover. \emph{Nature Reviews Microbiology} \textbf{15}, 502--510
(2017).}

\leavevmode\hypertarget{ref-Agosta2010How}{}%
\CSLLeftMargin{2. }
\CSLRightInline{Agosta, S. J., Janz, N. \& Brooks, D. R. How specialists
can be generalists: Resolving the {`parasite paradox'} and implications
for emerging infectious disease. \emph{Zoologia (Curitiba)} \textbf{27},
151--162 (2010).}

\leavevmode\hypertarget{ref-Power2004Pathogen}{}%
\CSLLeftMargin{3. }
\CSLRightInline{Power, A. G. \& Mitchell, C. E. Pathogen Spillover in
Disease Epidemics. \emph{The American Naturalist} \textbf{164}, S79--S89
(2004).}

\leavevmode\hypertarget{ref-Geoghegan2017Comparative}{}%
\CSLLeftMargin{4. }
\CSLRightInline{Geoghegan, J. L., Duchêne, S. \& Holmes, E. C.
Comparative analysis estimates the relative frequencies of co-divergence
and cross-species transmission within viral families. \emph{PLOS
Pathogens} \textbf{13}, e1006215 (2017).}

\leavevmode\hypertarget{ref-Muylaert2022Present}{}%
\CSLLeftMargin{5. }
\CSLRightInline{Muylaert, R. L. \emph{et al.} Present and future
distribution of bat hosts of sarbecoviruses: Implications for
conservation and public health. \emph{Proceedings of the Royal Society
B: Biological Sciences} \textbf{289}, 20220397 (2022).}

\leavevmode\hypertarget{ref-Ka-WaiHui2006Reasons}{}%
\CSLLeftMargin{6. }
\CSLRightInline{Ka-Wai Hui, E. Reasons for the increase in emerging and
re-emerging viral infectious diseases. \emph{Microbes and Infection}
\textbf{8}, 905--916 (2006).}

\leavevmode\hypertarget{ref-Mollentze2020Viral}{}%
\CSLLeftMargin{7. }
\CSLRightInline{Mollentze, N. \& Streicker, D. G. Viral zoonotic risk is
homogenous among taxonomic orders of mammalian and avian reservoir
hosts. \emph{Proceedings of the National Academy of Sciences}
\textbf{117}, 9423--9430 (2020).}

\leavevmode\hypertarget{ref-Latinne2020Origin}{}%
\CSLLeftMargin{8. }
\CSLRightInline{Latinne, A. \emph{et al.} Origin and cross-species
transmission of bat coronaviruses in China. \emph{bioRxiv: The Preprint
Server for Biology} 2020.05.31.116061 (2020)
doi:\href{https://doi.org/10.1101/2020.05.31.116061}{10.1101/2020.05.31.116061}.}
